---
title: "Edition of the 'initiatives of population cohorts' dataset for manual processing"
date:  "10/12/2021"
output:
  html_document:
    toc:           yes
    toc_float:     yes
    df_print:      paged
    code_folding:  hide
    code_download: yes
params:
  write_file: no
---

```{r setup, message=FALSE, warning=FALSE}
# Libraries:

library(knitr)
library(tidyverse)
library(readxl)
library(xlsx)
library(xfun)
library(rmarkdown)


# Constants:
ROOT_DIR <- ".."

# Output configuration:

opts_knit$set(root.dir = ROOT_DIR)
opts_chunk$set(message = FALSE)
```

```{r includes, cache=FALSE}
source("R/Constants.R", encoding = 'UTF-8')
```

# Introduction

The complete initiatives dataset requires some manual processing,
as it was agreed to strip down the free text fields in the table of initiatives
in order to reduce its contents and length (see
[minutes of 31/03/2022 meeting](output/meetings/2022-03-31_SYNCHROS_Mapping_initiatives_paper_editorial_meeting.md)
).
Some manual processing needs to be done to produce a shorter version of those
fields.
The table is re-written with duplicate columns for the fields to rewrite.

# Dataset loading

## Dataset with the table of initiatives

The table in the sheet ``r COMPLETE_TABLE_SHEET`` of file
``r INITIATIVES_FILENAME`` is first loaded.

```{r load-table, message=FALSE}
initiatives <- read_excel(INITIATIVES_FILEPATH, sheet = COMPLETE_TABLE_SHEET)
```

# Data processing

## Column addition

```{r free-text-cols}
free_text_cols_out <- paste(BULLET_PREFIX, FREE_TEXT_COLS, collapse = LINE_FEED)
```


The following are free text columns that are printed in the output,
and thus need to be stripped down, creating a shorter version:

`r free_text_cols_out`

We duplicate these columns, adding the duplicate versions to "strip down":

```{r}
initiatives <- initiatives |> mutate(
  across(
    all_of(FREE_TEXT_COLS),
    list(~.) |> set_names(FT_CONDENSED_SUFFIX)
  )
)
```

# Output

## Dataset saving

This throughput dataset is saved again in Excel format in order to send it to
the coauthors for the manual processing, and then use it in the next steps.

```{r save-dataset, eval=params$write_file}
# Set to `eval=FALSE` to avoid overwriting the result when executing
#   automatically (if necesssary, "Knit with Parameters...")

# Try to remove sheet (if exists):
if (file.exists(INIT_MANUAL_FILEPATH)) {
  
  excel_file <- loadWorkbook(INIT_MANUAL_FILEPATH)
  sheets     <- getSheets(excel_file) |> names()
  
  if (TABLE_1_SHEET %in% sheets) {
    
    excel_file |> removeSheet(sheetName = TABLE_1_SHEET)
  }
  excel_file |> saveWorkbook(file = INIT_MANUAL_FILEPATH)
}

initiatives       |>
  as.data.frame() |> # Necessary to avoid an error when writting a tibble
  write.xlsx(
    INIT_MANUAL_FILEPATH,
    sheet     = TABLE_1_SHEET,
    row.names = FALSE,
    showNA    = FALSE,
    append    = TRUE
  )
```

## Document writing

The output is also written into a .docx document, in order to facilitate
the edition.

```{r write-document, results='hide'}
Rscript_call(
  render,
  list(input = "throughput/Initiatives_free_text_fields.Rmd")
)
```
