---
title: "Additional analyses addressing reviewer 2 comments"
output:
  html_document:
    toc:          yes
    toc_float:    yes
    code_folding: hide
    df_print:     paged
---



```{r, include=FALSE}
source("src/Analyses.R")
```


# Topics by funding

Comment by reviewer 2:

> What type of data/ area of research was funded by private institutions
vs public ones? Were they different?


```{r, results='asis'}
tab1_new                                  |>
  select(funding, starts_with("topic"))   |>
  filter(!is.na(funding))                 |>
  mutate(funding = funding |> fct_drop()) |>
  tbl_summary(by = funding)
```


# Nº of cohorts (total & harmonized), participants (total & harmonized), and harmonized variables for initiatives with/without omics data

Comment by reviewer 2:

> I suspect that the cohorts with a large number of variables (ex 37k)
and large N must be cohorts related to cohorts with biological data (omics).

```{r}
tab1_new                |>
  filter(!is.na(omics)) |>
  select(
    omics,
    starts_with(c("cohorts.", "participants.", "harmonizedvariables"))
  )                     |>
  tbl_summary(
    by = omics,
    type = everything() ~ "continuous"
  )
```

<br>

We suspect the teams carrying out the initiatives _with omics data_ may still
be active.
To investigate this, we look at the frequency of this variable for intiatives
with/without omics data.

```{r}
tab1_new                |>
  filter(!is.na(omics)) |>
  tbl_summary(by = omics, include = team_activity)
```

