---
title: "Update of the 'initiatives of population cohorts' dataset"
date:  "10/12/2021"
output:
  html_document:
    toc:           yes
    toc_float:     yes
    df_print:      paged
    code_folding:  hide
    code_download: yes
---

```{r setup, message=FALSE, warning=FALSE}
# Libraries:

library(knitr)
library(tidyverse)
library(readxl)
library(vctrs)
library(micar)
library(stringr)
library(rlang)
library(glue)
library(countrycode)
library(magrittr)
library(xlsx)


# Constants:
ROOT_DIR <- ".."

# Output configuration:

opts_knit$set(root.dir = ROOT_DIR)
opts_chunk$set(message = FALSE)
```

```{r includes}
source("R/Constants.R", encoding = 'UTF-8')
```

# Introduction

This dataset is intended to update the list of initiatives that resulted from
the mapping reported in Syncrhos deliverable 1.3 with the latest information
included in the Syncrhos repository.
The aim is to include all the updated information in the initiatives in the list
in the manuscript about the mapping of initiatives
integrating population cohorts.


# Loading and formatting datasets

The list of initiatives in Table 1 in Laura's draft are copied and pasted
to the file *`r INITIATIVES_FILEPATH`* in Excel format,
putting it in the sheet *`r TABLE_1_SHEET`*.
Before copying it, it is necessary to do the following two procedures manually:

- Change the Windows local configuration so that the "comma" is considered as
  the thousands separator and the "stop" as the decimal separator
  (Control panel -> Region -> Additional configuration,
  then type `.` in "Decimal symbol", and `,` in "Thousands separation symbol",
  then click on "Apply").
  
- Change all the thousand separation symbols by "commas" manually in the Word
  document, then copy the table and paste it in Excel.

Without these two operations, it is possible that numbers are missinterpreted
in Excel and there are errors parsing them as numbers.

After manually performing these operations, I read the dataset.

```{r load-table-1-data, message=FALSE}
tab1_data <- read_excel(
  INITIATIVES_FILEPATH,
  sheet = TABLE_1_SHEET,
  col_types = "text",
  na    = c("NA", "na"),
  skip  = 1
)
```


The initiatives from the Mica repository are are accessed using
the API provided by Yannick.

**A warning for reproducibility**:
The `mica` R package does not allow to access a "screenshot version" of the
repository, i.e., to use a date parameter to indicate the repository the state
in which we would like to query the repository.
Therefore, if the repository changes because new information is added to the
initiatives, or new initiatives are published, this notebook will not be
reproducible.
Hence, the next chunk is set to `eval=FALSE`,
because it is not intended to be run for reproducing the results.

```{r get-mica-data, eval=FALSE}
# Date of query is 10/12/2021
repo_connection <- mica.login(url = SYNCHROS_URL)

mica_data <- mica.networks(repo_connection, limit = 100) |> as_tibble()

mica.logout(repo_connection)

mica_data |> write_csv(file = MICA_DATA_FILEPATH)
```


Instead of running the previous chunk, use the file `r MICA_DATA_FILEPATH`,
which has the results of the repository query saved as a `.csv` file
and is read in the next chunk.
Note that the column types are set to the same types as in the original
export from the repository (with the only exception of `dataEntryDate`,
which is directly transformed into a date vector).

```{r load-mica-data}
mica_data <- read_csv(
  MICA_DATA_FILEPATH,
  col_types = cols(
    .default                              = col_character(),
    studies                               = col_integer(),
    variables                             = col_integer(),
    collectedDatasets                     = col_integer(),
    collectedVariables                    = col_integer(),
    harmonizedDatasets                    = col_integer(),
    dataschemaVariables                   = col_integer(),
    yearCreated                           = col_integer(),
    nbParticipants.total                  = col_integer(),
    nbParticipants.harmonized             = col_integer(),
    dataEntryDate                         = col_date(format = ""),
    age.min                               = col_integer(),
    age.max                               = col_integer(),
    methodology.moreCohortsToBeHarmonized = col_logical(),
    methodology.nbHarmonizedVariables     = col_integer(),
    methodology.nbCohorts.total           = col_integer(),
    methodology.nbCohorts.harmonized      = col_integer()
  )
)
```


## Data wrangling of Table 1 from Laura's draft

Table 1 has a complex header, as well as several collapsed cells,
which makes it difficult to transform into a rectangular format.
I change the headers in the first place.

```{r table1-col-names, message=FALSE}
# Get the names from the first two rows of the file, and create names and labels
tab1_data_header <- read_excel(
  INITIATIVES_FILEPATH,
  sheet = TABLE_1_SHEET,
  n_max = 1
)                                                       |>
  pivot_longer(everything())                            |>
  mutate(
    name = name                              |>
      str_detect(AUTO_VARNAME_PREFIX_REGEXP) |>
      if_else(name |> dplyr::lag(1), name)
  )                                                     |>
  unite("label", name, value, sep = ": ", na.rm = TRUE) |>
  mutate(name = label |> vec_as_names(repair = "universal"))


# Rename the table with the new names

rename_vector <- tab1_data |>
  colnames()               |>
  set_names(tab1_data_header |> pull(name))

tab1_data <- tab1_data |> rename(!!!rename_vector)
```


Then I drop the empty rows and collapse cells that represent
a single initiative.
By visual inspection, it seems that each initiative has
a first, "main" row without any missing value,
while all the remaining rows have eight or more missing values.
(There is one exception: One initiative has a single missing value in
the column `Briefly.describe.the.initiative.s.population`,
because the population description starts in the second row; therefore,
I consider that main rows are the ones having either 0 or 1 missing value.)
I use those rows as a guide to extract the information from the rest of the rows
and create a single row per initiative.

```{r table1-collapse-rows}
n_cols <- tab1_data |> ncol()

# Compute the nº of missing values per row and identify the main rows
tab1_data <- tab1_data |>
  rowwise()            |>
  mutate(
    n_missing = sum(is.na(c_across())),
    main_row  = n_missing %in% 0:1
  )

# Drop completely empty rows
tab1_data <- tab1_data |>
  ungroup()            |>
  filter(n_missing < n_cols)

# Create a grouping variable using `main_row`
tab1_data <- tab1_data |>
  ungroup()            |>
  mutate(row = main_row |> cumsum())

# Group rows by initiative, then collapse all non-missing rows
#   in each initiative into a single row
tab1_data <- tab1_data                                                    |>
  ungroup()                                                               |>
  mutate(across(where(is.character), replace_na, replace = EMPTY_STRING)) |>
  group_by(row)                                                           |>
  summarize(across(where(is.character), glue_collapse, sep = SPACE))      |>
  select(-row)

# Assign meaningful labels (after summarizing to avoid losing them)
tab1_data <- tab1_data |> map2_dfc(
  tab1_data_header |> pull(label), # Labels to assign to the columns
  \(column, label) { column %@% "label" <- label; column }
)
```


## Column selection from the MICA dataset

After visual exploration,
I try to select from the MICA dataset what appear to be the columns in Table 1.

```{r select-mica-columns}
mica_data <- mica_data |>
  select(-harmonizedDatasets) # All values are 0

mica_data_tab1 <- mica_data |> select(
  acronym,
  name,
  region,
  countries,
  description,
  methodology.nbCohorts.total,
  methodology.nbCohorts.harmonized,
  methodology.moreCohortsToBeHarmonized,
  nbParticipants.total,
  nbParticipants.harmonized,
  age.min,
  age.max,
  
  methodology.nbHarmonizedVariables,
  cohortCriteria
)
```


# Acronym match of the two datasets

I first extract the initiative acronym from Table 1.
Some initiatives have the acronym in parentheses
(and they are different from other phrases in parentheses in that there are
no whitespaces), so I use that criterion first;
when it does not find the acronym,
I cut the string at the first whitespace in column `INITIATIVE`
(I also include colon `:`, as some initiatives have it right after the acronym),
and finally a few acronyms are manually edited to correct them and either
make them match the value in the MICA dataset.

After checking [Angel's paper](https://doi.org/10.3389/fpubh.2021.666844),
I identify a few patient and clinical trial initiatives from the repository
that should be filtered out.
Desiree identifies a few more in an email
(**Topic:** RE: Synchros -
Publications Mapping of initiatives that integrate European ... (UAM: Daniel),
on 23/11/2021 14:15),
that were mismatched, were also patient cohorts, or had a different acronym.
According to this information, only the initiatives already in Table 1 need to
be matched with the corresponding entries from the repository.

```{r tab1-mica-matching}
tab1_data <- tab1_data |>
  mutate(
    acronym = INITIATIVE                               |>
      str_extract("(?<=\\()[^\\s]+(?=\\))")            |>
      coalesce(INITIATIVE |> str_extract("^[^\\s:]+")) |>
      na_if("The")                                     |> # Values matched by
      na_if("Genome")                                  |> #   regexp that are
      na_if("Tohoku")                                  |> #   not actual
      na_if("EU")                                      |> #   acronyms
      recode( # Values to recode manually to match the MICA dataset acronyms
        EUCAN              = "EUCAN-Connect",
        `BBMRI-NL-Biobank` = "BBMRI-NL",
        Interconnect       = "interconnect",
        `CPTP-HP`          = "CanPath",
        g2aging            = "Gateway"
      ),
    id = acronym |> coalesce(INITIATIVE) # Fill in missing values with name
  )                    |>
  select(id, acronym, everything())
```

The `id` is created as an ancillary variable that contains the acronym,
except that when it is missing, it is filled in with the initiative name.

The repository entries then are filtered using the entries in the original
Table 1.

```{r update-matches}
mica_data_tab1 <- mica_data_tab1 |>
  mutate(id = acronym)           |> # Used to match with Table 1 later on
  select(id, everything())       |>
  semi_join(tab1_data,  by = "id")
```


# Construction of the updated dataset

The repository columns have different names from the ones in Table 1,
so the columns need to be match correspondingly between both datasets first.
I first explore the column names to see if they match each other

```{r column-names}
mica_data_tab1 |> colnames()
tab1_data      |> colnames()
```


## Formatting missing values

In the original Table 1, values "No information obtained" are recasted into
missing values

```{r recode-missing}
tab1_data <- tab1_data |> mutate(
  across(everything(), str_remove, pattern = NO_INFO_REGEXP),
  across(everything(), na_if,      y       = EMPTY_STRING)
)
```


## Regions and countries

The repository has one more column, because in Table 1 there is one single
column for "Region" and "Countries", which is in two different columns in the
repository.

```{r explore-regions-countries}
tab1_countries_regions  <- tab1_data |> select(id, acronym, Region..Country)
mica_countries_regions  <- mica_data_tab1 |>
  rowid_to_column(var = "row")            |> # Rows in the repository
  select(acronym, region, countries, row)

full_join(tab1_countries_regions, mica_countries_regions, by = "acronym")
```


The comparison among these columns show that the two share approximately the
same information, although the one in the repository is considered more up
to date.

```{r region-initiatives}
REGION_INITIATIVES <- c(
  "ICC-dementia",
  "interconnect",
  "InterLACE",
  "OMEGA-NET",
  "RAND",
  "Genome Asia 100K",
  "DEMETRIQ",
  "EU Child Cohort Network",
  "RECAP",
  "MORGAM",
  "NCI", # Should be dropped??
  "CHARGE"
)

region_initiatives_out <- paste(
  BULLET_PREFIX,
  REGION_INITIATIVES,
  collapse = LINE_FEED
)
```


The information in the repository is not 100% complete though,
so it is completed with the information already present in Table 1;
this mainly refers to the countries the cohort populations are taken from,
with the following exceptions:

`r region_initiatives_out`

Therefore, we fill in the `region` and `countries` columns from
the repository table with the appropriate values taken from the original
Table 1.
Before doing this, we must format the `countries` column
from the repository to transform the ISO codes into country names in English.
The content of `countries` and `region` is also standardized
by changing semicolons (`;`) and "and" conjunctions by commas.

```{r complete-regions-countries}
# Turn ISO3 codes into country names and collapse them
mica_data_tab1 <- mica_data_tab1 |>
  separate_rows(countries)       |>
  mutate(
    countries = countries |>
      countrycode(origin = 'iso3c', destination = 'country.name')
  )                              |>
  group_by(across(-countries))   |>
  summarize(
    countries   = countries |> glue_collapse(sep = COMMA),
    n_countries = n(), # Nº of countries per initiative, for descriptive stats
    .groups     = "drop"
  )

# Standardize separating countries and regions by commas
tab1_countries_regions <- tab1_countries_regions |>
  mutate(
    across(
      Region..Country,
      str_replace,
      pattern     = REPLACE_SEPS_REGEXP,
      replacement = COMMA
    )
  )

# Complete missing rows in the repository table
mica_data_complete <- mica_data_tab1 |>
  right_join(tab1_countries_regions, by = c("id", "acronym"))

# Fill in countries and regions with values in Table 1
mica_data_complete <- mica_data_complete |>
  mutate(
    countries = if_else(
      is.na(countries) & !(id %in% REGION_INITIATIVES),
      Region..Country,
      as.character(countries)
    ),
    region = if_else(
      is.na(region) & (id %in% REGION_INITIATIVES),
      Region..Country,
      region
    )
  ) |>
  select(-Region..Country)
```


## Other variables

The remaining columns are check side by side for correspondence.

```{r variable-correspondence}
# Column names correspondences between the two datasets
names_correspondence <- tibble(
  repository = mica_data_tab1                 |>
    select(-region, -countries, -n_countries) |>
    colnames(),
  table1     = tab1_data     |>
    select(-Region..Country) |>
    colnames()
) |>
  slice(-(1:2)) # Drop `id` and `acronym` from the columns to compare
  
# Variable correspondences
variable_correspondence_checks <- names_correspondence %$%
  map2(
    repository, table1,
    ~{
      full_join(
        mica_data_tab1 |> select(id, .data[[.x]]),
        tab1_data      |> select(id, .data[[.y]]),
        by = "id"
      )
    }
  )

# Correspondence of yes/no with logical values
cohorts_foreseen_check <- variable_correspondence_checks |>
  extract2(5)                                            |>
  group_by_at(-1)                                        |>
  count()
```


The results are not explicitly shown, but one can see that the columns generally
have the same information (sometimes rephrased), with one exception:
Instead of an "Age range" column, the repository has two columns,
for the minimum and maximum age, and it lacks the column corresponding to
"Is it a representative sample?").

Therefore, excluding those columns, we fill in the missing information in 
the repository table with the information in Table 1
(i.e. the information from the repository has a higher priority).
Regarding the column "Is it a representative sample?",
it has been decided to drop it from the table
(see email from Ellen on 15/12/2021, 10:24,
**subject:**
"RE: Quick question about the data exported from the Synchros repo".)

```{r fill-in-vars}
# Drop non-coincident columns
names_correspondence <- names_correspondence |>
  filter(repository |> str_detect("^age.", negate = TRUE))

# Iterate through the two dataset columns to fill in the missing information:
names_correspondence %$%
  walk2(
    repository, table1,
    ~{
      # Convert the character values to the same type as in the repository:
      
      cast_fun <- mica_data_complete |> pull(.x) |> typeof() |> switch(
        integer = as.integer,
        logical = \(x) {
          
          x           |>
            tolower() |>
            trimws()  %>%
            { case_when(. == "no" ~ FALSE, . == "yes" ~ TRUE, TRUE ~ NA) }
        },
        as.character # Default value, casts as `character` (i.e. does nothing)
      )
      
      suppressWarnings( # Suppresses warning by coercion of character to integer
        tab1_data <- tab1_data |> mutate(across(all_of(.y), cast_fun))
      )
      
      # Fill in the missing values where corresponding:
      mica_data_complete <<- mica_data_complete |>
        full_join(tab1_data |> select(id, .data[[.y]]), by = "id") |>
        mutate(
          !!sym(.x) := coalesce(.data[[.x]], .data[[.y]])
        ) |>
        select(-.data[[.y]])
    }
  )
```


## Initiative names

Some of the values in the `INITIATIVE` field include the acronym and the
initiative name in different formats.
We try to identify these and extract the name without the acronym.
The formats identified are:

- Initiative name with the acronym in parentheses at the end or in the middle.
- Acronym followed by initiative name in parentheses.
- Acronym followed by volon and initiative name.

```{r}
mica_data_complete <- mica_data_complete |> mutate(
  name = name                               |>
    str_remove(glue("\\s?\\({acronym}\\)")) |>
    str_remove(glue("{acronym}: "))         |>
    str_remove(glue("{acronym}\\s?\\("))    |>
    str_remove("\\)")
)
```


## Age range

The column "Age range of the sample" has the additional complication that
the information is recorded in the repository as two columns,
for the minimum and maximum age of the cohorts.
This information must be processed from the Table 1 first
(the values are visually checked and some exceptions manually corrected),
then the missing values filled in in the two columns of the complete dataset,
and then the "age range" column regenerated.

```{r fill-in-age-range}
# The general rule applied is: The first number to appear is the minimum age,
#   and the second one (the next after deleting the first one) is the maximum.
tab1_age <- tab1_data       |>
  transmute(
    id,
    Age.range.of.the.sample, # Included for visual exploration
    min              = Age.range.of.the.sample |> str_extract(NUMBER_REGEXP),
    age_range_min_rm = Age.range.of.the.sample |> str_remove(NUMBER_REGEXP),
    max              = age_range_min_rm        |> str_extract(NUMBER_REGEXP)
  )                         |>
  select(-age_range_min_rm) |>
  mutate(across(starts_with("m"), as.integer))

# The following exceptions are identified by visual exploration of `tab1_age`:
EXCEPTION_IDS <- tribble(
  ~id,           ~value,
   "ATHLOS",      "max",
   "LIFEPATH",    "min",
   "CHANCES",     "max",
   "NCI",         "max",
   "GenomEUtwin", "max" # Because it says "85+"
)

# Discard the original age range column (needed for visual inspection only)
tab1_age <- tab1_age |> select(-Age.range.of.the.sample)

# Exceptions are overwitten with missing values:
EXCEPTION_IDS %$% walk2(
  id, value,
  ~{
    tab1_age <<- tab1_age |> mutate(
      !!sym(.y) := id |> equals(.x) |> if_else(NA_integer_, .data[[.y]])
    )
  }
)

# The table is completed with the values computed just above:
mica_data_complete <- mica_data_complete |>
  full_join(tab1_age, by = "id")         |>
  mutate(
    age.min   = age.min |> coalesce(min),
    age.max   = age.max |> coalesce(max),
    age_range = glue("{age.min} - {age.max}") |>
      na_if('NA - NA')                        |>
      str_replace(' - NA', '+')
  )                                      |>
  select(-min, -max)
```


## Case deletion

```{r nonpopulation-cohorts}
NONPOP_COHORTS_INITIATIVES <- c("NCI")

nci_age_range <- tab1_data                 |>
  filter(id == NONPOP_COHORTS_INITIATIVES) |>
  pull(Age.range.of.the.sample)
```


According to the manuscript, only initiatives that integrate population cohorts
are considered, as the initiatives that integrate population cohorts and
patient cohorts / clinical trials are reported in Angel's paper.
Initiative `r NONPOP_COHORTS_INITIATIVES`
reports considering patient and clinical trial cohorts,
according to field "Age range of the sample":

    `r nci_age_range`

This initiative should therefore be dropped.

```{r drop-excluded-initiatives}
mica_data_complete <- mica_data_complete |>
  filter(id != NONPOP_COHORTS_INITIATIVES)
```


# Dataset saving

The throughput dataset resulting from this data wrangling process is saved
in Excel format in order to share it and use it in the next steps.

```{r save-dataset, eval=FALSE}
# Set to `eval=FALSE` to avoid overwriting the result when executing
#   automatically (change if necesssary)
mica_data_complete |>
  as.data.frame()  |> # Necessary to avoid an error when writting a tibble
  write.xlsx(
    INITIATIVES_FILEPATH,
    sheet     = UPDATED_TABLE_SHEET,
    row.names = FALSE,
    showNA    = FALSE,
    append    = TRUE
  )
```
