---
title: "Update of the 'initiatives of population cohorts' dataset"
date:  "10/12/2021"
output:
  html_document:
    df_print: paged
    code_folding: show
editor_options:
  chunk_output_type: console
---

```{r setup, message=FALSE}
# Libraries:

library(knitr)
library(tidyverse)
library(readxl)
library(micar)
library(stringr)
library(rlang)
library(glue)


# Constants:

INITIATIVES_FILENAME <- "synchros-initiatives.xlsx"
TABLE_1_SHEET        <- "Table 1"
MICA_DATA_FILENAME   <- "mica-export.csv"

SYNCHROS_URL <- "https://repository.synchros.eu"

ROOT_DIR <- ".."
DATA_DIR <- "dat"
INITIATIVES_FILEPATH <- file.path(DATA_DIR, INITIATIVES_FILENAME)
MICA_DATA_FILEPATH   <- file.path(DATA_DIR, MICA_DATA_FILENAME)

AUTO_VARNAME_PREFIX_REGEXP <- r"(^\.\.\.)"
EMPTY_STRING               <- ''
SPACE                      <- ' '

PATIENT_COHORTS_AND_CCTT <- c(
  "ART-CC",
  "B-CAST",
  "CNODES",
  "COSMIC",
  "High-Risk Myocardial Infarction Database Initiative", 
  "Lifebrain",
  "Sentinel Initiative",
  "STROKOG"
)

# Output configuraction:

opts_knit$set(root.dir = ROOT_DIR)
opts_chunk$set(message = FALSE)
```


# Introduction

This dataset is intended to update the list of initiatives that resulted from
the mapping reported in Syncrhos deliverable 1.3 with the latest information
included in the Syncrhos repository.
The aim is to include all the updated information in the initiatives in the list
in the manuscript about the mapping of initiatives
integrating population cohorts.


# Loading datasets

The list of initiatives in Table 1 in Laura's draft are copied and pasted
to the file *`r INITIATIVES_FILEPATH`* in Excel format,
putting it in the sheet *`r TABLE_1_SHEET`*.
Then I read the dataset.

```{r load-table-1-data}
tab1_data <- read_excel(
  INITIATIVES_FILEPATH,
  sheet = TABLE_1_SHEET,
  na    = c("NA", "na"),
  skip  = 1
)
```


The initiatives from the Mica repository are are accessed using
the API provided by Yannick.

**A warning for reproducibility**:
The `mica` R package does not allow to access a "screenshot version" of the
repository, i.e., to use a date parameter to indicate the repository the state
in which we would like to query the repository.
Therefore, if the repository changes because new information is added to the
initiatives, or new initiatives are published, this notebook will not be
reproducible.
Hence, the next chunk is set to `eval=FALSE`,
because it is not intended to be run for reproducing the results.

```{r get-mica-data, eval=FALSE}
# Date of query is 10/12/2021
repo_connection <- mica.login(url = SYNCHROS_URL)

mica_data <- mica.networks(repo_connection, limit = 100) |> as_tibble()

mica.logout(repo_connection)

mica_data |> write_csv(file = MICA_DATA_FILEPATH)
```


Instead of running the previous chunk, use the file `r MICA_DATA_FILEPATH`,
which has the results of the repository query saved as a `.csv` file
and is read in the next chunk.
Note that the column types are set to the same types as in the original
export from the repository (with the only exception of `dataEntryDate`,
which is directly transformed into a date vector).

```{r load-mica-data}
mica_data <- read_csv(
  MICA_DATA_FILEPATH,
  col_types = cols(
    .default                              = col_character(),
    studies                               = col_integer(),
    variables                             = col_integer(),
    collectedDatasets                     = col_integer(),
    collectedVariables                    = col_integer(),
    harmonizedDatasets                    = col_integer(),
    dataschemaVariables                   = col_integer(),
    yearCreated                           = col_integer(),
    nbParticipants.total                  = col_integer(),
    nbParticipants.harmonized             = col_integer(),
    dataEntryDate                         = col_date(format = ""),
    age.min                               = col_integer(),
    age.max                               = col_integer(),
    methodology.moreCohortsToBeHarmonized = col_logical(),
    methodology.nbHarmonizedVariables     = col_integer(),
    methodology.nbCohorts.total           = col_integer(),
    methodology.nbCohorts.harmonized      = col_integer()
  )
)
```


# Data wrangling of Table 1 from Laura's draft

Table 1 has a complex header, as well as several collapsed cells,
which makes it difficult to transform into a rectangular format.
I change the headers in the first place.

```{r table1-col-names}
# Get the names from the first two rows of the file, and create names and labels
tab1_data_header <- read_excel(
  INITIATIVES_FILEPATH,
  sheet = TABLE_1_SHEET,
  n_max = 1
) |>
  pivot_longer(everything()) |>
  mutate(
    name = name |>
      str_detect(AUTO_VARNAME_PREFIX_REGEXP) |>
      if_else(name |> dplyr::lag(1), name)
  ) |>
  unite("label", name, value, sep = ": ", na.rm = TRUE) |>
  mutate(name = label |> vctrs::vec_as_names(repair = "universal"))

# Rename the table with the new names
rename_vector <- tab1_data |>
  colnames() |>
  set_names(tab1_data_header |> pull(name))

# Assign meaningful labels
tab1_data <- tab1_data |>
  rename(!!!rename_vector) |>
  map2_dfc(
    tab1_data_header |> pull(label), # Labels to assign to the columns
    \(column, label) { column %@% "label" <- label; column }
  )
```


Then I drop the empty rows and collapse cells that represent
a single initiative.
By visual inspection, it seems that each initiative has
a first, "main" row without any missing value,
while all the remaining rows have eight or more missing values.
(There is one exception: One initiative has a single missing value in
the column `Briefly.describe.the.initiative.s.population`,
because the population description starts in the second row; therefore,
I consider that main rows are the ones having either 0 or 1 missing value.)
I use those rows as a guide to extract the information from the rest of the rows
and create a single row per initiative.

```{r table1-collapse-rows}
n_cols <- tab1_data |> ncol()

# Compute the nº of missing values per row and identify the main rows
tab1_data <- tab1_data |>
  rowwise() |>
  mutate(
    n_missing = sum(is.na(c_across())),
    main_row  = n_missing %in% 0:1
  )

# Drop completely empty rows
tab1_data <- tab1_data |>
  ungroup() |>
  filter(n_missing < n_cols)

# Create a grouping variable using `main_row`
tab1_data <- tab1_data |>
  ungroup() |>
  mutate(row = main_row |> cumsum())

# Group rows by initiative, then collapse all non-missing rows
#   in each initiative into a single row
tab1_data <- tab1_data |>
  ungroup() |>
  mutate(across(where(is.character), replace_na, replace = EMPTY_STRING)) |>
  group_by(row) |>
  summarize(across(where(is.character), glue_collapse, sep = SPACE))
```


# Column selection from the MICA dataset

After visual exploration,
I try to select from the MICA dataset what appear to be the columns in Table 1.

```{r select-mica-columns}
mica_data <- mica_data |>
  select(-harmonizedDatasets) # All values are 0

mica_data_tab1 <- mica_data |> select(
  acronym,
  name,
  region,
  countries,
  description,
  methodology.nbCohorts.total,
  methodology.nbCohorts.harmonized,
  methodology.moreCohortsToBeHarmonized,
  nbParticipants.total,
  nbParticipants.harmonized,
  age.min,
  age.max,
  methodology.nbHarmonizedVariables,
  cohortCriteria
)
```


# Comparison of both datasets

The first thing that calls one's attention is the difference in the number
of initiatives in each datsaet.

```{r n-rows}
n_tab1 <- tab1_data |> nrow()
n_mica <- mica_data_tab1 |> nrow()
```


While Table 1 in Laura's draft has `r n_tab1` initiatives,
the table exported from the MICA repository has `r n_mica`.

These results are inconsistent with what is stated in the section "Results"
of the draft.
The second paragraph states there are 59 initiatives,
close to the number in Table 1 but not the same.
Figure 1 also says there are 59 initiatives.

In order to find out which initiatives match in both tables,
I first extract the initiative acronym from Table 1.
Some initiatives have the acronym in parentheses
(and they are different from other phrases in parentheses in that there are
no whitespaces), so I use that criterion first;
when it does not find the acronym,
I cut the string at the first whitespace in column `INITIATIVE`
(I also include colon `:`, as some initiatives have it right after the acronym),
and finally a few acronyms are manually edited to correct them and either
make them match the value in the MICA dataset,
or fill them in with the name of the initiative
(when there is no acronym available).

```{r tab1-mica-matching}
tab1_data <- tab1_data |>
  mutate(
    acronym = INITIATIVE |>
      str_extract("(?<=\\()[^\\s]+(?=\\))") |>
      coalesce(INITIATIVE |> str_extract("^[^\\s:]+")) |>
      recode( # Values to recode manually to match the MICA dataset acronyms
        EUCAN              = "EUCAN-Connect",
        `BBMRI-NL-Biobank` = "BBMRI-NL",
        Interconnect       = "interconnect"
      ) |>
      na_if("The")    |> # Values matched by regexp that are not real acronyms
      na_if("Genome") |>
      na_if("Tohoku") |>
      na_if("EU")     |>
      coalesce(INITIATIVE) # Fill in the previous ones with initiative name
  )

tab1_data_join <- tab1_data |> select(acronym, row_tab1 = row)

mica_data_tab1_join <- mica_data_tab1 |>
  rownames_to_column("row") |>
  select(row_mica = row, acronym)

tab1_mica_matched <- full_join(
  tab1_data_join, mica_data_tab1_join,
  by = "acronym"
)
```


After checking [Angel's paper](https://doi.org/10.3389/fpubh.2021.666844),
I identify a few patient and clinical trial initiatives from the repository
that should be filtered out.
Desiree identifies a few more in an email
(**Topic:** RE: Synchros -
Publications Mapping of initiatives that integrate European ... (UAM: Daniel),
on 23/11/2021 14:15),
that were mismatched, were also patient cohorts, or had a different acronym.
According to this information I updated the matches
and filter out the initiatives that do not correspond
(i.e. initiatives that include patient cohorts and/or clinical trials)

```{r update-matches}
tab1_data <- tab1_data |>
  mutate(
    acronym = acronym |>
      recode( # Values to recode manually to match the MICA dataset acronyms
        `CPTP-HP`          = "CanPath",
        g2aging            = "Gateway"
      )
  )

tab1_data_join <- tab1_data |> select(acronym, row_tab1 = row)

mica_data_tab1_join <- mica_data_tab1 |>
  rownames_to_column("row") |>
  select(row_mica = row, acronym) |>
  filter(!acronym %in% PATIENT_COHORTS_AND_CCTT)

tab1_mica_matched <- full_join(
  tab1_data_join, mica_data_tab1_join,
  by = "acronym"
)
```


The resulting matches are all in Laura's Table 1.

```{r tab1-mica-matching-output}
tab1_mica_matched
```
