---
title: |
  Worldwide mapping of initiatives that integrate population cohorts
output:
  officedown::rdocx_document:
    base_format:     bookdown::word_document2
    reference_docx:  ../../www/IJE_template.docx
    fig_width:       6.73
    fig_asp:          .75
    number_sections: no
    tables:
      layout:        autofit
      caption:
        sep: '. '
    page_margins:
      bottom: 1
      top:    1
      right:  1
      left:   1
      gutter: 0
    keep_md:         no
bibliography:        ../../www/Mapping_initiatives.bib
csl:                 ../../www/sage-vancouver-brackets.csl
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# Environment setup:
gc()

# Libraries:
library(knitr)
library(flextable)

# Constants:
DOC_DIR      <- getwd()
ROOT_DIR     <- "../.."
SRC_DIR      <- "src"
SRC_FILEPATH <- file.path(SRC_DIR, "Analyses.R")

# Knitr configuration:
opts_knit$set(root.dir = ROOT_DIR) # Interferes with officedown!
opts_chunk$set(
  echo       = FALSE,
  results    = 'hide',
  message    = FALSE,
  warning    = FALSE,
  cache      = TRUE,
  autodep    = TRUE,
  dpi        = 300,
  fig.retina = 1,
  tab.topcaption = TRUE,
  dev.args   = list(png = list(type = "cairo"))
)

# Output formatting options:
set_flextable_defaults(
  font.family       = "Times New Roman",
  padding.top       = 1,
  padding.bottom    = 0
)
```

```{r load-chunks, cache=FALSE}
SRC_FILEPATH |> read_chunk()
```

```{r includes, cache=FALSE}
```

```{r load-data}
```

```{r country-derivate-vars}
```

```{r health-topic-derivate-vars}
```

```{r socioenvcontext-derivate-vars}
```

```{r socioenvcontext-other-values}
```

```{r analyses-derivate-vars}
```

```{r create-output-dataset}
```

```{r preprocess-descriptives}
```

```{r create-headers-labels}
```

```{r summary-table}
```

```{r descriptives-table}
```

```{r compute-descriptive-values}
```

```{r uncache-descriptive-values, cache=FALSE, include=FALSE}
# Total nº of initiatives:
total_N

# Maximum number of missing values in one descriptive variable:
max_missing_out

# Active projects:
active_projects_prop_out
active_projects_prop_total_out

# Countries of leading institutions:
country_institution_max_freq_out
country_institution_ranking_next_out

# Funding sources:
funding_public_out
funding_mixed_out

# Harmonization strategy:
harmo_strategy_max_out
harmo_max_prop_out

# Analysis:
analysis_pooled_out
analysis_meta_out
analysis_federated_out

# Omics:
omics_out

# Total nº of cohorts:
total_cohorts_range_out
total_cohorts_median_out

# Harmonized cohorts:
harm_cohorts_max_out

# More cohorts to be harmonized:
more_cohorts_expected_n_out

# Harmonized variables:
harm_vars_range_out
harm_vars_median_out

# Harmonized participants:
harm_participants_max_out
harm_participants_median_out

# Particinats' ages across included cohorts:
max_age_max_out

# Topics:
prop_ageing_out

# Nº of countries:
max_n_countries_out
min_n_countries_out

# Nº of continents
prop_1_continent_out
prop_6_continents_out

# Nº of initiatives per region
prop_Europe_out

n_Africa_out
prop_Africa_out

n_Latam_out
prop_Latam_out
```

<br>

Laura Alejandra Rico-Uribe^1,2^;
Ángel Rodríguez-Laso^3^;
Ellen Vorstenbosch^1,4^;
Andreas J. Weser^5^;
Laura Fincias^4^;
Yannick Marcon^6^;
Leocadio Rodriguez-Mañas^3^;
Josep María Haro^4,7^;
José Luis Ayuso-Mateos^1,8,9,\*^;
Daniel Morillo-Cuadrado^1,9^

<br>

^1^ CIBERSAM (Network-based Biomedical Research Consortium,
area of Mental Health), Instituto de Salud Carlos III, Spain

^2^ School of Health Sciences, Universidad Internacional de La Rioja

^3^ CIBERFES (Network-based Biomedical Research Consortium,
    area of Frailty and Healthy Ageing), Instituto de Salud Carlos III, Spain

^4^ Parc Sanitari Sant Joan de Déu, Barcelona, Spain

^5^ HUNT Research Centre, Norwegian University of Science and Technology,
    Levanger, Norway

^6^ Epigeny<!-- # TODO: Complete Yannick's affiliation details -->

^7^ Universitat de Barcelona, Barcelona, Spain

^8^ Department of Psychiatry, Universidad Autónoma de Madrid, Madrid, Spain

^9^ Instituto de Investigación Sanitaria La Princesa (IIS-LP),
    Hospital Universitario de La Princesa, Madrid, Spain

^\*^ **Corresponding author**:
José Luis Ayuso-Mateos,
Universidad Autónoma de Madrid, Department of Psychiatry.
Arzobispo Morcillo 4, 28029 Madrid, Spain. E-mail: joseluis.ayuso@uam.es

**Word count:** <!-- # TODO: Fill in (CURRENT: 2708) -->

\newpage

# Abstract

**Background:** Health research increasingly requires the analysis of
data across different cohorts;
statistical power and representativeness are paramount to study
infrequent conditions or specific groups of individuals.
Although integrating cohorts is often difficult,
several initiatives have done so.
A map portraying a complete overview of these efforts is missing.
This paper aims to lay out the methodology and results of
such mapping, focusing on initiatives that integrate population cohorts.  
**Methods:** Three different sources were used to search for initiatives:
1) A systematic search in the MEDLINE database;
2) suggestions of potential initiatives to be included,
provided by the partners of the SYNCHROS consortium who devised the mapping,
and European Commission scientific officers; and
3) descendent searches in the references and links of the initiatives obtained
from the two previous sources.  
**Results:** We found `r total_N` initiatives that
integrate population cohorts, although a significant proportion of
their leading teams were inactive or unresponsive.
Topics, socio-environmental contexts, and populations were
rather disparate, with high volumes of integrated data.
Most of the initiatives were led by European and North American institutions,
and most of the integrated cohorts came from these two areas as well.  
**Conclusions:** The number of cohort integration initiatives is still low,
especially compared with the large amount of cohort studies
being carried out worldwide.
The existing and new initiatives and cohort studies should strive toward
improving transparency and availability of information.
We expect this mapping exercise to help researchers
integrate cohorts and find venues for collaborative research.

**Keywords:** mapping; population cohorts; harmonisation; integration

\newpage

# Key messages

* A mapping of the initiatives across the globe that integrate
  population cohorts is carried out.  

* The number of initiatives found is rather low compared with
  individual cohort studies being conducted.

* Integrating population cohorts may help increase
  statistical power and representativeness in epidemiological research.

\newpage

# Background

There has been a recent surge in scientific data collection
in the healthcare domain [@roski_creating_2014].
Several large-scale cohort studies have recently been or are being
conducted nowadays around the globe:
the Millennium Cohort Study [@connelly_cohort_2014],
the Environmental Health Risks in European Birth Cohorts
[@nieuwenhuijsen_environmental_2011],
the Collaborative Research on Ageing in Europe
[@leonardi_collaborative_2013],
or the English Longitudinal Study of Ageing [@steptoe_cohort_2013]
are popular examples.
Integrating these individual cohorts is the next step to getting
the most valuable information out of them.
Cohort integration studies may increase sample sizes drastically,
allowing the study of important but infrequent phenomena
(e.g. rare diseases) [@thompson_thinking_2009] while avoiding publication bias.
The pressure is rapidly building up to keep up-to-date with
integrating and capitalizing on this information,
and several initiatives have joint forces for integrating cohorts
from different studies
(e.g. ATHLOS @sanchez-niubo_data_2017; BioSHaRE @kaye_consent_2016;
CHICOS @strandberg-larsen_association_2017; HELIX @maitre_human_2018).
However, methodological and infrastructure issues [@panahiazar_empowering_2014],
as well as concerns about legal (i.e., data protection) and ethical aspects
impose constraints that may hinder the potential benefits of
these integration studies.
An overview of what is being done to overcome these challenges is still missing
in the scientific landscape.
Systematizing this knowledge may help further integrate existing cohorts,
as well as streamline the design and integration of future cohorts.

In an attempt to expand and make these efforts more systematic,
the European Commission called for a sustainable, strategic agenda for
a better, global coordination of cohorts.
The project
*SYNergies for Cohorts in Health: integrating the ROle of all Stakeholders*
(SYNCHROS) was funded by the Horizon 2020 Research and Innovation Program with
the goal of formulating this strategy,
through intensive stakeholder collaboration (https://www.synchros.eu).
<!-- # TODO: Ref. needed?. -->
Among others, one of its first actions was the mapping of
the cohort integration initiatives landscape
across the world and in Europe especially.
The objective was to obtain first-hand information about
the methodologies and solutions implemented for integrating patient,
clinical-trial, and population cohorts.

The patient cohort landscape mapping has already been reported elsewhere
[@rodriguez-laso_map_2021].
This paper focuses on the state of the art in
the integration of population cohorts.
We present the methodology for gathering a collection of initiatives
that have integrated such cohorts in the last twenty years,
as well as the results and conclusions of this mapping exercise.

# Methods

## Identification of initiatives

Three different methods were used
to find initiatives that integrate population cohorts.
The first one was a systematic search in the MEDLINE database.
The second one consisted of asking the SYNCHROS consortium and
scientific officers of the European Commission for suggestions on initiatives.
The third was a descendent search using the information
(references, descriptions, and links) from the two previous sources.

## Database search in MEDLINE

The search was intended to produce a representative,
albeit non-exhaustive, list of cohort integration initiatives.
An initial set of search terms were agreed upon by the consortium partners.
In order to accomplish a manageable number of hits, these terms were tested and
the ones that yielded more than 500 hits were discarded.
(To reduce the number of hits, the term "cohort" was added
to some searches before the discarding.)
New terms were extracted from abstracts of relevant retrieved papers
and tested again.
To obtain the most recent scientific evidence,
the search was limited to studies published in 2000 or later.
The search was conducted in July 2019.
The final search query was:

```
(cohort OR
  “prospective study” OR
  “longitudinal study” OR
  "individual meta-analysis"[All Fields] OR
  "individual participant data meta-analysis"[All Fields] OR
  "individual patient data meta-analysis"[All Fields] OR
  "individual meta analysis"[All Fields] OR
  "individual participant data meta analysis"[All Fields] OR
  "individual patient data meta analysis"[All Fields] OR
  "meta analysis using individual"[All Fields] OR
  "meta-analysis using individual"[All Fields] OR
  "meta analysis of individual"[All Fields] OR
  "meta-analysis of individual"[All Fields] OR
  "mega-analysis"[All Fields] OR
  "mega analysis"[All Fields])  
AND  
("harmonization study" OR
  "integration study" OR
  "integration initiative" OR
  "integrated study" OR
  "merged cohort" OR
  "data pooling" OR
  "pooled sample" OR
  "combined data" OR
  "combining data" OR
  "harmonized data" OR
  "harmonised data" OR
  "harmonizing data" OR
  "data harmonization" OR
  "data harmonisation" OR
  "data sharing" OR
  "common database" OR
  "multiple cohorts" OR
  "multiple longitudinal studies" OR
  "international consortium" OR
  "collaborative effort").  
AND  
("2000/01/01"[Date - Publication] : "2019/07/31"[Date - Publication])  
AND  
English[Language]  
AND  
Humans[MeSH]
```

<br>

## Selection of initiatives

**Inclusion criteria.** Initiatives published in English from 2000 to July 2019
were included if they integrated health population cohorts of any age
(birth, adolescents, adults, elderly, oldest old),
and included sociodemographic, lifestyle, biological, genetic, omics
(genomics, proteomics, metabolomics), imaging, or environment factors data.
Additionally, at least one of the integrated cohorts had to
have information about the sample at two or more time points.  
**Exclusion criteria.** Initiatives without available data about (or access to)
their descriptive information
(webpage, a main report describing the main aim(s) in detail) were excluded.
Initiatives that integrated clinical trial cohorts and/or patient cohorts
are the subject of another publication [@rodriguez-laso_map_2021]
and thus were also excluded.

For all the initiatives found, a double-check was performed by
two different researchers in order to carry out an objective evaluation and
reduce the risk of bias.
In case of discrepancy between the two reviewers, a third person was consulted.

## Data extraction and analysis

The following technical information was extracted from the initiatives:
name, principal investigator (PI), initiative partners,
name of the leading institution,
contact person, information source, whether the research team
was active at the time of consultation, main objectives,
criteria for the cohorts to be included,
funding resources (`r FUNDING_LABELS_PRINT`),
and a brief description of the population addressed.
When the information was not available through the website or
published articles,
the principal investigator and/or the contact person of the project
was consulted, by email first and, if no reply was received,
by telephone or postal mail.

Some variables were collected in order to describe the harmonisation process,
namely: harmonisation strategy
(prospective/retrospective ex-ante/retrospective ex-post),
number of harmonised cohorts,
whether more cohorts were foreseen to be harmonised,
number of participants with harmonised data,
maximum number of harmonised variables
(including those where harmonisation was not possible for all the cohorts),
setting of the harmonised cohorts (local-regional/national/international),
The following variables were deemed relevant for the integration effort and
thus were also collected for each initiative:
total number of cohorts, total number of participants, age range of the sample,
country/ies included, whether metadata and individual data were accessible to
other researchers,
whether any of the cohorts included biological (omics) samples,
research topic(s) and socio-environmental context(s) addressed,
and type(s) of aggregated analyses performed on the integrated cohorts
(pooled, federated, or meta-analysis).

The country where the leading institution was based and the countries
included in the integrated cohorts were recoded into continents
(`r VECTOR_CONTINENTS_AMERICAS_SEP_PRINT`).
For all the variables, descriptive statistics were computed when appropriate:
absolute frequencies and proportions (over number of non-missing values)
for the categorical variables;
median, minimum and maximum for the quantitative ones.
In the case of the number of participants with harmonised data and
the maximum number of harmonised variables, the statistics were computed
using only the initiatives with harmonised cohorts.

# Results

The flowchart diagram in Figure \@ref(fig:flowchart) summarizes the initiatives
included and excluded using the three methods.
The MEDLINE search identified 843 articles that described initiatives
to be potentially included. After screening the titles and abstracts,
166 articles met the inclusion criteria.
From the full-text review, 155 articles were excluded due to one or more of
the following reasons: one-time efforts (articles where the merging of data
was done only for carrying out the research presented in the paper),
initiatives already submitted by partners, articles with
cross-sectional harmonisation only, multisite cohorts/harmonisation of waves,
meta-analyses, reviews, case-control studies, patients' cohorts,
clinical trial cohorts, and others.
From the systematic review, 11 articles were included.
Additionally, from the partners’ and officers' suggestions and
the descendent searches,
189 potential initiatives were obtained. After screening for eligibility,
141 projects were excluded for one or more of the following reasons:
initiatives published before the year 2000,
projects that did not integrate cohorts,
one initiative with only cross-sectional cohorts,
and one initiative that integrated patient and clinical trial cohorts.
This resulted in the inclusion of another 48 initiatives that integrate
population cohorts.
A total of `r total_N` initiatives were included;
they are presented in Supplementary Table S1,
along with a subset of their most representative information.
This information is explained in Table \@ref(tab:summary),
along with the data from *`r sample_initiative`* shown as an example.

It is worth noting that in a large number of cases the contact person was
unreachable throughout the whole mapping process,
so the proportion of missing data was as large as
`r max_missing_out` for some variables.
At the time of consultation, `r active_projects_prop_out` of
the research teams that had carried out the projects remained active
(although this percentage drops to `r active_projects_prop_total_out` if
we assume that those which did not respond were inactive).
The majority of the leading institutions were based in
`r country_institution_max_freq_out`, followed by institutions
from `r country_institution_ranking_next_out`.
`r funding_public_out` of the initiatives were funded by public institutions,
while `r funding_mixed_out` received funding from
both private and public institutions.

The total number of cohorts ranged between `r total_cohorts_range_out`
(median = `r total_cohorts_median_out`).
Some initiatives had up to `r harm_cohorts_max_out` harmonised cohorts,
while others only had collected the participating cohorts
(but had not started harmonising them yet).
At least `r more_cohorts_expected_n_out` initiatives planned
to obtain and harmonise more cohorts in the future.
The harmonised cohorts had a large variability in the maximum number of
harmonised variables, ranging from `r harm_vars_range_out`
(median = `r harm_vars_median_out`).
They would comprise up to `r harm_participants_max_out` participants,
although the median value was around `r harm_participants_median_out`.
`r harmo_strategy_max_out` (retrospective) harmonisation was
by far the most prevalent strategy,
used by `r harmo_max_prop_out` of the initiatives.
The majority of initiatives (`r analysis_pooled_out`) used pooled analysis
to perform integrated analysis on the cohorts. Meta-analysis and federated
analysis were less common, being used only by `r analysis_meta_out` and
`r analysis_federated_out` of the initiatives
(note that some initiatives used more than one type of integrated analysis).

The targeted populations ranged in age from childbirth to
`r max_age_max_out` years old
(although this probably refers to an unbounded upper limit).
Ageing and air pollution were the most prevalent topic and
socio-environmental context, respectively,
each one addressed by `r prop_ageing_out` of the initiatives, and
`r omics_out` of the initiatives included omics data (i.e. biological samples).
The cohorts were collected in as many as `r max_n_countries_out` countries,
although there were also initiatives with cohorts from
only `r min_n_countries_out` country.
Most of them (`r prop_1_continent_out`) were circumscribed
to just one continent,
being Europe the most frequent (`r prop_Europe_out`),
while only `r prop_6_continents_out` comprised cohorts from all
six continents.
Latin America and Africa were the most underrepresented continents,
with only `r n_Latam_out` (`r prop_Latam_out`)
and `r n_Africa_out` (`r prop_Africa_out`)
initiatives including cohorts from them.
The complete descriptive information of the initiatives can be found in Table
\@ref(tab:descriptives).
Also, the most up-to-date information on each initiative can be consulted
in the repository of the SYNCHROS project (https://repository.synchros.eu).

# Discussion

The purpose of this mapping was to gather knowledge about
the state of the art of the initiatives that integrate population cohorts.
The landscape of these initiatives has been revealed to be quite disparate,
with a high variability of populations and variables,
as well as topics and regions addressed by the integrated cohorts.
As foreseeable, the number of participants and variables integrated is larger
in general than in initiatives integrating patient cohorts
[@rodriguez-laso_map_2021].
Taking the high volumes of data in some of the initiatives,
their potential for future researchers is undeniable.
Still, the number of integration initiatives seems rather low,
compared to the overall and increasing number of very large cohort datasets
worldwide.<!-- # TODO: Reference needed? -->
It is also worth stressing that a large proportion of the initiatives found are
inactive nowadays, limiting the possibilities of data access and sharing.

Most of the initiatives found were funded by public entities
(and mainly by the European Commission).
We found a scarcity of private funding,
with only a few initiatives being partially funded by private institutions.
One might think that e.g. pharmaceutical companies would be highly interested in
integrating evidence across cohorts and thrive towards personalized medicine,
especially in patient and clinical trial cohorts.
However, the participation of private entities in funding patient cohorts was
even lower [@rodriguez-laso_map_2021].
Nevertheless, note that these funding sources refer only to the integration
initiatives;
this information does not take into account other possible funding sources for
the individual cohort studies integrated into an initiative,
so private interest in funding cohort research might be overlooked.

Focus on environmental exposures was more prevalent than
genetic and biological factors.
Very few of the initiatives were actually found to harmonise and integrate
biological samples
(although there was a high rate of missingness in this variable).
Including them may help discover possible causal pathways among biological,
behavioural, social, demographic, economic, and health outcomes
[@hobcraft_reflections_2009].
Unfortunately, incorporating bio-measures in cohort research is not always
feasible.

Most of the initiatives were led by European and American institutions;
as would be expected, most of the integrated cohorts were also collected in
these two continents.
Interestingly, very few of the initiatives included cohorts focused
on African and Latin American & Caribbean countries.
Previous studies emphasize the relative lack of health studies conducted in
low- and middle-income countries (LMICs @lawlor_birth_2009).
More representativeness of LMICs would be necessary to grant
the external validity of the cohort studies that inform
global health policy recommendations.

## Strengths and weaknesses

To our knowledge, this is the first effort to map and describe in detail all
the initiatives integrating population cohorts.
Given the difficulties reported by authors when integrating cohorts
[@gatz_data_2015],
we expect positive outcomes of our endeavour in three main aspects.
First, the SYNCHROS repository is a resource where interested researchers can
find integrated population cohort data or contribute their population cohorts.
Second, several PIs and project managers have provided first-hand information
on the barriers and solutions they have found when integrating cohorts.
Finally, we expect all this information to be extremely relevant to designing
a European strategy for cohort integration,
such as is the aim of the SYNCHROS project.

Although we must raise awareness of the non-exhaustive nature of this mapping,
we should stress that this manuscript aims at representativeness
rather than exhaustiveness.
It should be noted though that, given the huge amount of work involved in
integrating population cohorts, we deem it unlikely that we have missed
any initiatives that have published results.
Also, as the response rate of the principal investigators was rather low,
relevant information on some initiatives is still missing
at the time of submission of this manuscript.
However, the SYNCHROS repository is an evolving project,
where the most recent information available is continuously updated.

## Conclusions

Knowledge about restricted populations and phenomena
(e.g. personalized medicine, rare diseases, epigenetics)
require massive sample sizes to achieve the necessary
statistical power.
Moreover, global representativeness can only be achieved by addressing
more diverse populations [@henrich_most_2010]
from different ethnic, regional or/and socioeconomic settings.
Synergies across a wide variety of existing cohort integration projects
would help pursuing these goals, while being more cost-effective than
undertaking new international mega-cohorts [@larsen_pregnancy_2013].

However, there is still a shortfall of initiatives that integrate 
worldwide population cohorts.
Most of the currently available cohorts probably lack the necessary
transparency and availability of information
(variables, study designs, data access, etc.)
to afford multi-study research.
We expect the SYNCHROS repository to be a valuable resource for
emerging collaborative research,
with a spotlight in developing and enriching a "learning healthcare system"
[@frohlich_hype_2018].

# Ethics approval

<!-- # TODO: Complete -->

# Author contributions

LR contributed to the acquisition, analysis and interpretation of data,
and wrote the first draft of the manuscript.
DM analysed and interpreted the data
and wrote the subsequent versions of the manuscript.
ARL contributed substantially to the conception and design of
the data collection and its acquisition,
contributed to the writing of the manuscript, and revised it critically.
EB and AW contributed to the writing of the manuscript and
its critical revision.
All authors approved the final version of the manuscript and
agree to be accountable for all aspects of the work.

# Data availability

<!-- # TODO: Complete -->

# Supplementary material

Supplementary material is available at IJE online

# Funding

This work was supported by the European Union’s Horizon 2020
research and innovation programme [grant number 825884],
and the Network-based Biomedical Research Consortium,
area of Mental Health (CIBERSAM), Instituto de Salud Carlos III
to LR and DM.

# Acknowledgments

We acknowledge and send our gratitude to the principal investigators and
contacts of the projects that answered all our communications.
We would also like to acknowledge Maelstrom Research for
their active contribution to the development of the Mica repository engine.

# Conflicts of interest

None declared.

# References

::: {#refs}
:::

\newpage

```{r set-flextable-wd, cache=FALSE}
# Necessary for flextable to find the csl file:
opts_knit$set(root.dir = DOC_DIR)
```

```{r summary, cache=FALSE, tab.id="summary", tab.cap="Summary of the main information shown in the table of initiatives.", results='asis'}
initiatives_summary_output
```

\newpage

```{r descriptives, cache=FALSE, tab.id="descriptives", tab.cap="Descriptive statistics of the initiatives mapped.", results='asis'}
descriptives_table_output
```

\newpage

```{r flowchart, cache=FALSE, fig.cap="Flowchart diagram detailing the initiatives included and excluded.", results='asis'}
include_graphics("../../www/Figure_1.tif", dpi = NA)
```

```{r reset-root-dir, cache=FALSE}
# Necessary for officedown to find the template file:
opts_knit$set(root.dir = ROOT_DIR)
```
